Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
edits section 3

Our approach to causal structure learning combines domain knowledge with data-driven
insights in a manner that is based on the following considerations: 
(1) A domain experts is possibly good at determining causal directions between variables if 
there is a clear causal direction between them. (2) A domain expert may be less good
at identifying cases where there is no causal relationship between the variables,
since \emph{potential} causal relationships can often be argued for anyway.
(3) Many domain experts could struggle to distinguish direct from indirect 
effects, since the presence of a direct effect between two variables
of interest depends on all other variables present in the graph.

@@

We therefore consider the following two models of a domain experts: Given two variables
$ X $ and $ Y $ that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(X \gets Y, X \to Y) & \textrm{otherwise} \\
\end{cases}
$$

That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect cases where there is no cause-effect
relationship between two variables. Using such oracles for DAG construction therefore 
inherently risks that superfluous edges are created. In our approach, such superfluous edges
will be detected and removed by utilizing the data.

To model how our learning procedure interacts with the data, we assume that we
additionally have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$. These are the standard oracles considered in constraint-based structure learning
algorithms, such as the PC algorithm \cite{Spirtes2001}.

Based on these two oracles, we now define two core procedures that we will use to 
iteratively change a current DAG structure. The procedure 
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses conditional independence
information to search for missing edges in the graph and then uses domain
knowledge to orient them. The procedure \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
conditional independence information to remove superfluous edges from the
graph.

@@

The function $\textsc{Expand}$ takes an initial list of edges and searches for 
any vertex pairs that are not connected but where the d-separation oracle 
indicates  that there is still a residual association not explained by the other variables.
An optional parameter $B$, the empty set by default, allows to specify a ``black list''
of vertex pairs that must not be connected, and another optional parameter $k$
specifies the maximum amount of edges to be added by this procedure.

Let $G^+$ denote the transitive closure of $G$, that is, the graph in which 
there is a direct edge $X \to Y$ whenever $X \in \textrm{An}_G(Y)$. The transitive
closure is related to the result of $\textsc{Expand}$ by the following two 
easily proved facts.

@@

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ just like when using a strong ancestral oracle; since only correctly oriented
edges are added between ancestor-descendant pairs, such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, an arbitrary 
edge is added in a manner that preserves acyclicity. The result is a supergraph
of $G^+$.
\end{proof}

@@

Since we require expert knowledge only in the $\textsc{Expand}$ operation, we may try
to be more economical by asking fewer questions at a time and interleaving expansion and
pruning steps. This leads us to the following, more incremental DAG construction algorithm.

@@

\begin{proof}
The loop in Algorithm~\ref{algo:expert} terminates if and only if $E=E^*$. If the loop does not 
terminate, a new edge has been added in line 6, and/or one or more edges were pruned in line 7; the pruned 
edges never include the edge added in line 6 in the same iteration.
Every edge can be added at 
most once and pruned at most once. Therefore, the algorithm always terminates
after at most $|V|(|V|-1)+1$ iterations of the loop. For every edge $e=X\to Y $ in the skeleton of $G$, $\mathcal{D}_G(X,Y,\mathbf{Z})=0$ irrespective of $\mathbf{Z}$, and $e$ must be added to $E$ in some iteration, and can never be pruned afterwards. Therefore, after some iteration, $(V,E)$ must be a supergraph of $G$ after executing line 6, and this will be pruned to the real graph $G$ in line 7 (Proposition~\ref{prop:prune}). In the next iteration, no further changes are made, and the loop terminates.
\end{proof}

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing each iteration of the \textsc{ExpertInLoop} algorithm. The dashed edges show all the potential edges that \textsc{Expand} step could add in each iteration.}
\end{figure}

@@

The \textsc{ExpertInLoop} algorithm adds a new edge in each iteration. The edge
to be added is selected using the \textsc{Expand} algorithm, which returns a
potential edge at random based on the iteration order. However, if the residual
association between the selected variables is low, adding the edge may result
in only a marginal improvement in the overall model fit. Given that the
algorithm requires expert intervention in each iteration to specify the edge
orientation, it is beneficial to prioritize edges that contribute the most to
improving the model.

To achieve this, instead of selecting edges randomly, we propose ranking
potential edges based on their residual association given the current DAG. This
residual association can be quantified using the effect size measures from the
CI tests used for querying the d-separation in Algorithm~\ref{algo:expand}.
Specifically, for a d-separation query $ D_G(X, Y, \mathbf{Z}) $, we quantify
the residual association, $ \phi(X, Y, \bm{Z}) $, as the effect size of the CI
test $ X \ci Y \rvert \bm{Z} $. In Algorithm~\ref{algo:ranked_expand}, we
outline a greedy approach to edge selection using the effect size measure.

To implement Algorithm~\ref{algo:ranked_expand}, we need to have an effect size
metric by which we can rank effects. Depending on the type of variables, we can
use the effect size measures shown in Section~\ref{sec:ci_tests}. The algorithm
selects the edge that has the highest absolute residual association, resulting
in prioritization of edges that contribute the most to improving the model.

\subsection{Connection to Score-Based Methods}

Our procedure shares similarities with score-based automated causal discovery
methods, such as Greedy Equivalence Search (GES), where the algorithm
iteratively adds or removes edges that maximize improvements in a specified
scoring metric. Given a scoring metric, --total residual association, $ \tau $ for 
a DAG $ G = (V, E) $, defined as:

@@

our \textsc{RankedExpand} approach behaves similarly to GES. In each iteration,
we add the edge between the variable pair that results in the lowest $ \tau $
and remove edges that have no impact on $ \tau $.

While both methods aim to make iterative modifications to the DAG, a key
distinction lies in the nature of the evaluation criteria: unlike standard
scoring metrics, $ \tau $ is not decomposable. Decomposibility is a key
property of scoring metrics, allowing them to be expressed as the sum of
individual variables in the model. This property ensures that adding or
removing an edge only affects a localized part of the model. In contrast, $
\tau $ can not be decomposed, as modifying an edge in one part of the model can
affect the residual associations in other parts of the model.

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ provides an absolute measure of model fit -- its value
approaches $ 0 $ as the model perfectly explains the observed data.

>>>>

Our approach to causal structure learning combines domain knowledge with data-driven
insights in a manner that is based on the following considerations: 
(1) A domain expert is possibly good at determining causal directions between variables if 
there \emph{is} a clear causal direction between them. (2) A domain expert may be less good
at identifying cases where there is no causal relationship between the variables,
since \emph{potential} causal relationships can often be argued for anyway.
(3) Many domain experts could struggle to distinguish direct from indirect 
effects, since the presence of a direct effect between two variables
of interest depends on all other variables present in the graph.

@@

We consider two models of a domain experts. Given two variables
$ X $ and $ Y $ that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(\}X \gets Y, X \to Y\}) & \textrm{otherwise} \\
\end{cases}
$$
where $\textrm{rand}(S)$ denotes a uniformly chosen element of $S$.
That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect the absence of cause-effect
relationships between two variables. Using such oracles for DAG construction  
inherently risks that superfluous edges are created.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \cite{Spirtes2001}.

Based on these two oracles, we now define two core procedures that
iteratively change a current DAG structure. The procedure 
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses CI  
information (supplied in this case by the d-separation oracle) 
to search for unexplained associations in the graph and 
then uses domain knowledge to determine ancestry. The 
procedure \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
CI information to remove superfluous edges from the
graph.

@@

$\textsc{Expand}$ takes an initial list of edges and searches for 
any unconnected vertex pairs that are not connected but where the
d-separation oracle indicates a residual association not explained 
by other paths.
The parameter $B$ specifies a ``black list''
of vertex pairs that must not be connected, 
whereas $k$
can be used to limit the maximum amount of edges 
to be added by this procedure; the usefulness of the $B$ and $k$
parameters will become clear soon.

Let $G^+$ denote the transitive closure of $G$ (the graph in which 
there is a direct edge $X \to Y$ whenever $X \in \textrm{An}_G(Y)$). 
The transitive closure is related to the result of $\textsc{Expand}$ 
by the following two easily proved facts.

@@

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected 
in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ like when using a strong ancestral oracle; 
since only correctly oriented
edges are added between ancestor-descendant pairs, 
such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, 
an arbitrary edge is added in a manner that preserves acyclicity.
The result is a supergraph of $G^+$.
\end{proof}

@@

Since we require expert knowledge only in the $\textsc{Expand}$ 
operation, we may try to be more economical by asking fewer 
questions at a time and interleaving expansion and
pruning steps. This leads us to the following, more iterative 
DAG construction algorithm.

@@

\begin{proof}
The loop in Algorithm~\ref{algo:expert} terminates if and 
only if $E=E^*$. If the loop does not  terminate, a new
 edge has been added in line 6, and/or one or more edges were 
pruned in line 7; the pruned edges never include the edge added 
in line 6 in the same iteration. Every edge can be added at 
most once and pruned at most once. Therefore, the algorithm 
always terminates after at most $|V|(|V|-1)+1$ iterations of
the loop. For every edge $e=X\to Y $ in the skeleton of $G$, 
$\mathcal{D}_G(X,Y,\mathbf{Z})=0$ irrespective of $\mathbf{Z}$, 
and $e$ must be added to $E$ in some iteration, and can never 
be pruned afterwards. Therefore, after some iteration, 
$(V,E)$ must be a supergraph of $G$ after executing line~6, 
and this will be pruned to the real graph $G$ in line~7 
(Proposition~\ref{prop:prune}). In the next iteration, 
no further changes are made, and the loop terminates.
\label{thm:itworks}
\end{proof}

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing each iteration of \textsc{ExpertInLoop}. The dashed edges show all the potential edges that \textsc{Expand} could add in each iteration.}
\end{figure}

@@

\textsc{ExpertInLoop} adds a new edge in each iteration. The edge
to be added is selected using the \textsc{Expand} algorithm, which 
returns a potential edge at random based on the iteration order. 
However, if the residual association between the selected variables 
is low, adding the edge may result in only a marginal improvement 
in the overall model fit. Given that the
algorithm requires expert intervention in each iteration to specify the edge
orientation, it can be beneficial to prioritize edges that contribute 
the most to improving the model.

To achieve this, instead of selecting edges randomly, we propose ranking
potential edges based on their residual association given the current DAG. This
residual association can be quantified using the effect sizes from the
CI tests used for deciding d-separation in Algorithm~\ref{algo:expand}.
Specifically, for a d-separation query $D_G(X, Y, \mathbf{Z}) $, we quantify
the residual association, $ \phi(X, Y, \bm{Z}) $, as the effect size of the CI
test $X \ci Y \rvert \bm{Z} $.

Algorithm~\ref{algo:ranked_expand} shows a 
version of {\sc Expand} where we use the largest unexplained association to
determine where to next add an edge. 
To implement this, we need an effect size
metric by which we can rank effects. Depending on the type of variables, 
we can use the effect size measures shown in Section~\ref{sec:ci_tests}. 
The algorithm selects the edge that has the highest absolute residual 
association, resulting
in prioritization of edges that contribute the most to improving the model.

\subsection{Comparison to Score-Based Methods}

Superficially, the data-driven aspects of our procedure appear
 similar to score-based automated causal discovery methods like 
Greedy Equivalence Search (GES), which 
iteratively add or remove edges that maximize improvements in a specified
scoring metric. Indeed, we can define the \emph{total residual association}
 $ \tau $ for a DAG $ G = (V, E) $ as:

@@

Our \textsc{RankedExpand} approach behaves similarly to GES in that it 
tries to prioritize modifications that lead to the largest improvements 
in $\tau$. However, adding an edge between the two vertices that contribute
the most to $\tau$ does not necessarily lead to the largest 
possible decrease in $\tau$ because the new edge can also affect other 
residual associations. In other words, $\tau$ is not a \emph{decomposable}
metric that can be expressed as a sum over individual variables in the model
and their parents.

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ could be seen as an absolute measure of model fit -- 
its value approaches $ 0 $ as the model perfectly explains the observed data.
